Staged Abstract Interpreters

### 2019-03-27 paper/40_absinterp.tex
section 5

\begin{lstlisting}
  def lift_nd[T](vs: List[T]): AnsM[T]
  def ap_clo(ev: EvalFun)(rator: Value, rand: Value): Ans = for {
    CloV(Lam(x, e), ρ: Env) <- lift_nd[AbsValue](f.toList)
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- ext_env(ev(e))(x → α)
  } yield rt
\end{lstlisting}

>>>>

\begin{lstlisting}
  def lift_nd[T](vs: List[T]): AnsM[T]
  def ap_clo(ev: EvalFun)(rator: Value, rand: Value): Ans = for {
    CloV(Lam(x, e), ρ: Env) <- lift_nd[AbsValue](rator.toList)
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- ext_env(ev(e))(x → α)
  } yield rt
\end{lstlisting}

### 2019-03-27 paper/50_stageabs.tex
section 5

\paragraph{Primitive Operations}

Once more, the way we handle closures is the same as in the staged concrete
interpreter: the recursive call to @ev@ with the body expression @e@ is compiled
and specialized, the wrapper function @f@ will be a field value in a
@CompiledClo@ object and be generated for the next stage. At the end, we return
a singleton set:

\begin{lstlisting}
  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(List[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ ⊔ Map(α → arg)
        ev(e)(ρ_*)(σ_*)(in)(out)
    }
    Set[AbsValue](emit_compiled_clo(f, λ, ρ))
  }
\end{lstlisting}

\begin{lstlisting}
  def lift_nd[T](vs: Rep[List[T]]): AnsM[T]
  def ap_clo(ev: EvalFun)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    clo <- lift_nd[AbsValue](rator.toList)
    σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val res: Rep[(List[(Value, Store)], Cache)] = emit_ap_clo(clo, rand, σ, in, out)
    _ <- put_out_cache(res._2)
    vs <- lift_nd[(Value, Store)](res._1)
    _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

Our fixed-point iteration again relies on two caches @in@ and @out@, but the
iteration no longer be done at the current stage. Because the @in@ and @out@ are
both next-stage values, the test of whether @in@ and @out@ are equal is a
generated expression in the next stage, and we can only know the comparison
result at the next stage. In other words, we do not know how many iterations we
need to reach the fixed-point. To achieve this, we need to stage a function
value --- @iter_aux@ is generated as a recursive function of type @Rep[Unit => (Value,Store)]@ 
and will be invoked at the next stage.

\begin{figure}[t!]
  \centering
\begin{lstlisting}
  def fix(ev: EvalFun => EvalFun): EvalFun = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val cfg: Rep[Config] = (unit(e), ρ, σ)
    val res: Rep[(List[(Value, Store)], Cache)] =
      if (out.contains(cfg)) (out(cfg).toList, out) // a next-stage if expression
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- ev(fix(ev))(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }
    _ <- put_out_cache(res._2); vs <- lift_nd(res._1); _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\vspace{-1em}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}

However, the instrumented evaluation function that uses the @in@ cache and
updates the @out@ cache can be completely eliminated by staging. Each recursive
call to @cached_ev@ will also be specialized if it is applied on subexpressions
of the analyzed program.

\begin{lstlisting}
\end{lstlisting}

\paragraph{Code Generation} Code generation

@@

Our staging schema works well in theory, but in practice would suffer from code
explosion, and possibly runtime GC overhead when analyzing (specializing) large
programs. In this section, we present several optimizations that largely
mitigate these issues. Still, implementing these optimizations do not need to
change the generic interpreter.

\paragraph{Specialized Data Structures} \todo{need revise}

Now we have already obtained an end-to-end staged abstract interpreter that is
able to specialize an analysis. However, we treat the data structures such as
@Map@s as black-boxes, which means any operations on a @Map@ become code in the
next stage. But, as we identified when introducing the generic interface, the
keys of any environment maps are identifiers in the program, which are
completely known statically. This leaves us a chance to further specialize the
data structures. Assume the @Map[K,V]@ is implemented as a hash map, if the keys
$K$ are known, then the indices can be computed statically. Thus the specialized
map would be an array @Array[Rep[V]]@ whose elements are next-stage values; all
the accesses to the array is determined during staging.

Particularly, if we are specializing a monovariant analysis, the address space
is equivalent to the identifiers, then the environment component can be entirely
eliminated, and the store is a specialized map as array of @Rep[Value]@
elements.

@@

\paragraph{Heterogeneous Staging} The generated program is in A-Normal form. \todo{TODO}

@@

@@

% direct-style vs CPS

>>>>

\paragraph{Primitive Operations} We keep our road map that focuses on functions
and applications. The @close@ method now is a rough mixing of the staged
concrete and unstaged abstract version: we have a current-stage function @f@,
that takes four next-stage values, @in@ and @out@ additionally; inside of @f@,
we collapse the @Ans@ monad to values of type @Result@ by providing the desired
arguments, i.e., the new environment, new store, and caches. The collapse
happens at current-stage, so the call of @ev@ is unfolded after staging. At
last, we generate a singleton set containing the compiled closure
@emit_compiled_clo(f)@, represented by an IR node in LMS.

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(List[(Value,Store)], Cache)]): Rep[AbsValue]
  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(List[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ ⊔ Map(α → arg)
        ev(e)(ρ_*)(σ_*)(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the set of closures into the monad stack, and generate a next-stage value
using @emit_ap_clo@, which represets the result of application, and finally we
put the @out@ cache, store and values back into the monads. The @emit_ap_clo@
method takes the target closure @clo@, argument @rand@, store and caches.

\begin{lstlisting}
  def lift_nd[T](vs: Rep[List[T]]): AnsM[T]
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(List[(Value, Store)], Cache)]
  def ap_clo(ev: EvalFun)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    clo <- lift_nd[AbsValue](rator.toList)
    σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val res: Rep[(List[(Value, Store)], Cache)] = emit_ap_clo(clo, rand, σ, in, out)
    _ <- put_out_cache(res._2)
    vs <- lift_nd[(Value, Store)](res._1)
    _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

Our fixed-point iteration again relies on two caches @in@ and @out@, which are
both staged values now. Therefore, the query of whether the @out@ cache contains
the current configuration produces a next-stage Boolean value, i.e.,
@Rep[Boolean]@. Consequently, the branching can not be determined statically --
we need to generate code for @if@. Figure \ref{fig:staged_coind_cache} shows the
staged version of @fix@.

\vspace{-1em}
\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: EvalFun => EvalFun): EvalFun = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val cfg: Rep[Config] = (unit(e), ρ, σ)
    val res: Rep[(List[(Value, Store)], Cache)] =
      if (out.contains(cfg)) (out(cfg).toList, out) // a next-stage if expression
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- ev(fix(ev))(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }
    _ <- put_out_cache(res._2); vs <- lift_nd(res._1); _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\vspace{-1em}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}
\vspace{-1em}

The @res@ variable is a next-stage result, which consists of a next-stage @if@
expression. The true branch simply returns a pair of queried result form the
@out@ cache and the @out@ cache. The else branch constructs a monad @m@ first,
which evaluates @e@ under the new @out@ cache. After which, we collapse the
monad @m@ with desired arguments. Finally, we obtain a placeholder @res@ since
it is a next-stage value, and put the content of @res@ back into the moand
stack.

\paragraph{Code Generation} The IR nodes for compiled closures and
applications are defined as following:

\begin{lstlisting}
  case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(List[(Value, Store)], Cache)]) extends Rep[AbsValue]
  case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                     in: Rep[Cache], out: Rep[Cache]) extends Rep[(List[(Value, Store)], Cache)]
\end{lstlisting}

The code generator for these two IR nodes is similar to staged concrete
interpreter, which generates a next-stage @CompiledClo@ object and a next-stage
function call to @f@, respectively.

@@

Our staging schema works well in theory, but in practice would suffer from code
explosion, and possibly runtime GC overhead (at the next stage) when analyzing
(specializing) large programs. In this section, we present several optimizations
that largely mitigate these issues. Still, implementing these optimizations do
not need to change the generic interpreter.

\paragraph{Specialized Data Structures}

Although every components except @Expr@s are staged, we treat the data
structures such as @Map@s as black-boxes, which means the operations on a @Map@
directly become code in the next stage, but we don't inspect any further inside.
As we identified when introducing the generic interface, the keys of @Env@s are
identifiers (i.e., strings) in the program, which are completely known
statically. This leaves us a chance to further specialize the data structures.
For example, let's assume that the @Map[K, V]@ is implemented as a hash map. If
the keys of type @K@ are all known statically, then the indices also can be
computed statically. Thus the specialized map would be an array of type
@Array[Rep[V]]@, whose elements are next-stage values; the size of the array is
also known statically. An access to the map is translated to an access to the
array with pre-determined index during staging.

Particularly, if we are specializing a monovariant analysis, the address space
is equivalent to the set of identifiers in the program. Then the accesses to the
environment can be completely determined statically, and generates the addresses
directly. After which, the stores can be specialized as arrays of @Rep[Value]@
elements.

@@

%\paragraph{Heterogeneous Staging} The generated program is in A-Normal form. \todo{TODO}

@@

\paragraph{What the generated code look like?} In the generated code, all the
primtive operations (such as @eval@, @fix@, @ap_clo@, etc.) and moandic
operations (such as @flatMap@ and @map@) are eliminated. The residual program
consists of statements and expressions that purely manipulate the environment,
store, and two caches, whose underlying representions are all @Rep[Map[K,V]]@. We
also have several operations on tuples and lists, which are residualized from
the use of monads internally.

@@
